The dynamics of starvation and recovery

### 2017-06-09 Manuscript/Nature-Submission/Nature-temp.tex
Nature submission

%\usepackage{scicite}
%\usepackage{times}
%\usepackage{graphicx}
%\usepackage{verbatim}
%\usepackage{subfigure}
%\usepackage{tikz}
%\usepackage{array}
%\usepackage{tabularx}
%\usepackage{multirow}
%\usepackage{multicol}
%\usepackage{multibox}
%\usepackage{rotating}
%\usepackage{layout}
%\usepackage{epstopdf}
%\usepackage{afterpage}
%\usepackage[left]{lineno}

%% PNAStmpl.tex
%% Template file to use for PNAS articles prepared in LaTeX
%% Version: Apr 14, 2008

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% BASIC CLASS FILE

\documentclass{pnastwo}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% OPTIONAL GRAPHICS STYLE FILE

\usepackage[pdftex]{graphicx}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% OPTIONAL POSTSCRIPT FONT FILES

\usepackage{pnastwoF}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% ADDITIONAL OPTIONAL STYLE FILES

%\usepackage{amssymb,amsfonts,amsmath}
\usepackage{amsmath}
\usepackage{amssymb}

% \usepackage[font=large]{caption}
\usepackage{subfigure}
\usepackage{epstopdf}
\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{array}
\usepackage{tabularx}
\usepackage{multirow}
\usepackage{color}
%\usepackage{multibox}
\usepackage{rotating}
\usepackage{color}
%\usepackage{lineno}
\usepackage[left]{lineno}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% OPTIONAL MACRO FILES
% \usepackage[comma,sort&compress]{natbib}
% \setlength{\abovedisplayskip}{3pt}
% \setlength{\belowdisplayskip}{3pt}
% 
% \usepackage[font=small,labelfont=bf]{caption}
% 
% \newcommand{\sid}[1]{\textcolor{red}{\bf [#1]}}

\bibliographystyle{pnas}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Don't type in anything in the following section:
%%%%%%%%%%%%
%% For PNAS Only:
\contributor{Submitted to Proceedings
of the National Academy of Sciences of the United States of America}
\url{www.pnas.org/cgi/doi/10.1073/pnas.0709640104}
\copyrightyear{2011}
\issuedate{Issue Date}
\volume{Volume}
\issuenumber{Issue Number}
%%%%%%%%%%%%

@@

\linenumbers
\setlength\linenumbersep{3pt}

%\title{Ecological and evolutionary implications of starvation and body size}

%\title{Survival of the Stable: Dynamic Consequences of Starvation and Reproduction}

%\title{Size Constraints on Starvation \& Reproduction Dynamics}

%\title{Eco-evolutionary Consequences of Starvation and Reproduction: Survival
%of the Stable}

\title{The dynamics of starvation and recovery}%: Eco-evolutionary feedbacks}

\author
{Justin D. Yeakel\affil{1}{School of Natural Science, University of California Merced, Merced, CA} \affil{2}{The Santa Fe Institute, Santa Fe, NM}\affil{4}{Contributed equally}\affil{5}{To whom correspondence should be addressed: jdyeakel@gmail.com}, Christopher P. Kempes \affil{2}{The Santa Fe Institute, Santa Fe, NM}\affil{4}{Contributed equally}, \and Sidney Redner \affil{2}{The Santa Fe Institute, Santa Fe, NM}\affil{3}{Department of Physics, Boston University, Boston MA}\affil{4}{Contributed equally}
}

\contributor{Submitted to Proceedings of the National Academy of Sciences
of the United States of America}

@@

\begin{article}

\begin{abstract} %250 words
The eco-evolutionary dynamics of species are fundamentally linked to the energetic constraints of its constituent individuals. Of particular importance are the tradeoffs between reproduction and the dynamics of starvation and recovery in resource-limited environments. To elucidate the consequences of this tradeoff, we introduce a minimal nutritional state-structured model that incorporates two classes of consumer: nutritionally replete consumers that reproduce, and undernourished, non-reproducing consumers that are susceptible to mortality. As a function of the transition rates between these two states that are determined by the abundance of resources, the consumer populations can either undergo cyclic dynamics or reach a steady state. We obtain strong constraints on starvation and recovery rates by deriving allometric scaling relationships between body size and a variety of traits and find that population dynamics subject to these constraints are typically driven to a steady state. Moreover, we find that these rates fall within a `refuge' in parameter space, where the probability of extinction of the consumer population is minimized. Thus we identify a potential mechanism that may both drive and constrain the dynamics of animal populations. Our model provides a natural framework that predicts maximum body size for mammals by determining the relative stability of an otherwise homogeneous population to a mutant population with altered percentage of body fat. For body masses $< 1.748\times10^7$g, individuals with increased energetic reserves can invade resident populations, and vice versa for body mass $> 1.748\times10^7$g, thus providing a principled mechanism for a within-lineage driver of Cope's rule.
\end{abstract}
\keywords{foraging | starvation|reproduction}

\noindent {\small {\bf Significance Statement} %105/120 words
Energetic investment in somatic maintenance and growth vs. reproduction directly impacts the dynamics of populations among species. Here, we construct a Nutritional State-structured Model (NSM) to assess the population-level effects of starvation and recovery of a consumer population in a resource-limited environment, and use allometric scaling relationships for mammals to establish all timescales and rates. Our model: i. reveals that mammalian energetic rates minimize the probability of stochastic extinction, ii. establishes dynamic bounds on mammalian body size while providing independent theoretical support for the energy equivalence hypothesis, and iii. provides a mechanistic driver for the evolutionary trend towards larger body size known as Cope's rule.}

\vspace{2mm}

@@

\begin{figure}
\centering
\includegraphics[width=0.4\textwidth]{fig_FixedPoint2.eps}
\caption{\small{ The transcritical (dashed) and Hopf bifurcation (solid) as a
  function of the starvation rate $\sigma$ and recovery rate $\rho$ for a 100g consumer.  These
  bifurcation conditions separate parameter space into unphysical, cyclic,
  and steady state dynamic regimes.  The colors show the steady state densities for the energetically replete consumers $F^*$, (warmer colors denote higher densities).  
  % Steady state densities for the energetically deficient consumers $H^*$ (not shown)
  % scale with those for $F^*$.  
}\label{fig:fp}}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.4\textwidth]{Growth-trajectory-diagram.eps}
\caption{\small{ The growth trajectory over absolute time of an individual organism as a function of body mass.  
Initial growth follows the black trajectory to an energetically replete reproductive adult mass $m=\epsilon_\lambda M$. %which we assume is 95\% asymptotic mass $M$.  
Starvation follows the red trajectory to $m = \epsilon_\sigma \epsilon_\lambda  M$. 
Recovery follows the green curve to the replete adult mass, where this trajectory differs from the original growth because only fat is being regrown which requires a longer time to reach $\epsilon_\lambda M$. %different energetics and 
Alternatively, death from starvation follows the blue trajectory to $m=\epsilon_\mu \epsilon_\lambda  M$.}\label{fig:growth}}
\end{figure}

@@

\begin{figure*}
\centering
\includegraphics[width=0.8\textwidth]{fig_ExtinctionAllometricComb3.eps}
\caption{\small{ Probability of extinction for a consumer with (\emph{A}) $M=10^2$g, (\emph{B}) $M=10^4$g, and (\emph{C}) $M=10^6$g as a function of the starvation rate $\sigma$ and recovery rate $\rho$, where the initial density is given as $(XF^*,XH^*,R^*)$, where $X$ is a random uniform variable in $[0,2]$. Note the change in scale for $M=10^4$ and $M=10^6$g.  Extinction is defined as the population trajectory falling below $0.2\times$ the allometrically constrained steady state. The white points denote the allometrically constrained starvation and recovery rate.}\label{fig:ext}}
\end{figure*}

@@

\begin{figure}
\centering
\includegraphics[width=0.48\textwidth]{fig_FPAllometric2.eps}
\caption{\small{ (\emph{A}) Consumer steady states $F^*$ (green) and $H^*$ (orange) as a function of
  body mass. Inset: Resource steady state $R^*$ as a function of consumer body mass.
  (\emph{B}) Total energetic use $B_{\rm tot}$ of consumer populations at the steady state as a function of body mass.
  The data are from Damuth \cite{Damuth:1987kr} and have been converted to total population metabolism using the allometric relationships for metabolic rate (please see SI and Refs. \cite{West:2001bv,hou,moses2008rmo}).}\label{fig:mass}}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.4\textwidth]{fig_Invasion.eps}
\caption{\small{ Invasion feasibility for organisms with a proportional change in
  mass $\chi$ against a population with a resident body mass $M$.  The blue
  region denotes proportions of modified mass $\chi$ resulting in successful invasion.  The
  red region denotes values of $\chi$ that result in a mass that is below the
  starvation threshold and are thus infeasible.
  Arrows point to the predicted optimal mass from our model $M_{\rm opt}=1.748\times 10^7$, which may serve as an evolutionary attractor for body mass.
  The black wedge points to the largest body mass known for terrestrial mammals (\emph{Deinotherium} spp.) at $1.74\times10^7$g~\cite{Smith:2010p3442}.}\label{fig:invasion}}
\end{figure}

@@

% \bibliography{aa_starving3}
\begin{thebibliography}{10}

\bibitem{Martin:1987dl}
Martin TE
\newblock (1987) {Food as a limit on breeding birds: A life-history
  perspective}.
\newblock \emph{Annu. Rev. Ecol. Syst.} 18:453--487.

\bibitem{Kirk:1997cc}
Kirk KL
\newblock (1997) {Life-history responses to variable environments: Starvation
  and reproduction in planktonic rotifers}.
\newblock \emph{Ecology} 78:434--441.

\bibitem{Kempes:2012hy}
Kempes CP, Dutkiewicz S, Follows MJ
\newblock (2012) {Growth, metabolic partitioning, and the size of
  microorganisms}.
\newblock \emph{Proc. Natl. Acad. Sci. USA} 109:495--500.

\bibitem{Mangel:1988uaa}
Mangel M, Clark CW
\newblock (1988) \emph{{Dynamic Modeling in Behavioral Ecology}}
\newblock (Princeton University Press, Princeton).

\bibitem{Mangel:2014kz}
Mangel M
\newblock (2014) {Stochastic dynamic programming illuminates the link between
  environment, physiology, and evolution}.
\newblock \emph{B. Math. Biol.} 77:857--877.

\bibitem{Yeakel:2013hi}
Yeakel JD, Dominy NJ, Koch PL, Mangel M
\newblock (2014) {Functional morphology, stable isotopes, and human evolution:
  a model of consilience}.
\newblock \emph{Evolution} 68:190--203.

\bibitem{Morris:1987eo}
Morris DW
\newblock (1987) {Optimal allocation of parental investment}.
\newblock \emph{Oikos} 49:332--339.

\bibitem{Tveraa:2003fq}
Tveraa T, Fauchald P, Henaug C, Yoccoz NG
\newblock (2003) {An examination of a compensatory relationship between food
  limitation and predation in semi-domestic reindeer}.
\newblock \emph{Oecologia} 137:370--376.

\bibitem{Daan:1988va}
Daan S, Dijkstra C, Drent R, Meijer T
\newblock (1989) in \emph{Acta XIX Congressus Internationalis Ornithologici,
  Volume I: Proceedings XIX International Ornithological Congress, 1986,
  Ottawa}, ed{} Ouellet H
\newblock (Proceedings XIX International Ornithological Congress, Ottawa), pp
  392--407.

\bibitem{Jacot:2009dv}
Jacot A, Valcu M, van Oers K, Kempenaers B
\newblock (2009) {Experimental nest site limitation affects reproductive
  strategies and parental investment in a hole-nesting passerine}.
\newblock \emph{Animal Behaviour} 77:1075--1083.

\bibitem{Stearns:1989ip}
Stearns SC
\newblock (1989) {Trade-offs in life-history evolution}.
\newblock \emph{Funct. Ecol.} 3:259.

\bibitem{Barboza:2002in}
Barboza P, Jorde D
\newblock (2002) {Intermittent fasting during winter and spring affects body
  composition and reproduction of a migratory duck}.
\newblock \emph{J Comp Physiol B} 172:419--434.

\bibitem{Threlkeld:1976ih}
Threlkeld ST
\newblock (1976) {Starvation and the size structure of zooplankton
  communities}.
\newblock \emph{Freshwater Biol.} 6:489--496.

\bibitem{Weber:1998jg}
Weber TP, Ens BJ, Houston AI
\newblock (1998) {Optimal avian migration: A dynamic model of fuel stores and
  site use}.
\newblock \emph{Evolutionary Ecology} 12:377--401.

\bibitem{Mduma:1999cp}
Mduma SAR, Sinclair ARE, Hilborn R
\newblock (1999) {Food regulates the Serengeti wildebeest: a 40-year record}.
\newblock \emph{J. Anim. Ecol.} 68:1101--1122.

\bibitem{Moore:2014hi}
Moore JW, Yeakel JD, Peard D, Lough J, Beere M
\newblock (2014) {Life-history diversity and its importance to population
  stability and persistence of a migratory fish: steelhead in two large North
  American watersheds}.
\newblock \emph{J. Anim. Ecol.} 83:1035--1046.

\bibitem{Mead:1989dt}
Mead RA
\newblock (1989) in \emph{Carnivore Behavior, Ecology, and Evolution}, ed{}
  Gittleman JL
\newblock (Springer US, Ithaca), pp 437--464.

\bibitem{Sandell:1990kw}
Sandell M
\newblock (1990) {The evolution of seasonal delayed implantation}.
\newblock \emph{Q Rev Biol} 65:23--42.

\bibitem{Bulik:1999eo}
Bulik CM, {et~al.}
\newblock (1999) {Fertility and reproduction in women with anorexia nervosa}.
\newblock \emph{J. Clin. Psychiat.} 60:130--135.

\bibitem{Trites:2003cc}
Trites AW, Donnelly CP
\newblock (2003) {The decline of Steller sea lions Eumetopias jubatus in
  Alaska: a review of the nutritional stress hypothesis}.
\newblock \emph{Mammal Rev.} 33:3--28.

\bibitem{Glazier:2009hq}
Glazier DS
\newblock (2009) {Metabolic level and size scaling of rates of respiration and
  growth in unicellular organisms}.
\newblock \emph{Funct. Ecol.} 23:963--968.

\bibitem{Kooi2000}
Kooijman SALM
\newblock (2000) \emph{{Dynamic Energy and Mass Budgets in Biological Systems}}
\newblock (Cambridge).

\bibitem{Sousa:2010ez}
Sousa T, Domingos T, Poggiale JC, Kooijman SALM
\newblock (2010) {Dynamic energy budget theory restores coherence in biology}.
\newblock \emph{Philos. T. Roy. Soc. B} 365:3413--3428.

\bibitem{Diekmann:2010da}
Diekmann O, Metz JAJ
\newblock (2010) {How to lift a model for individual behaviour to the
  population level?}
\newblock \emph{Philos. T. Roy. Soc. B} 365:3523--3530.

\bibitem{murdoch:2003}
Murdoch WW, Briggs CJ, Nisbet RM
\newblock (2003) \emph{{Consumer-resource Dynamics}}, Monographs in population
  biology
\newblock (Princeton University Press, Princeton) Vol.{}~36.

\bibitem{Benichou:2014wu}
Benichou O, Redner S
\newblock (2014) {Depletion-Controlled Starvation of a Diffusing Forager}.
\newblock \emph{arXiv} pp 1--5.

\bibitem{Benichou:2016wl}
B{\'e}nichou O, Chupeau M, Redner S
\newblock (2016) {Role of depletion on the dynamics of a diffusing forager}.
\newblock \emph{Journal of Physics A: {\ldots}}.

\bibitem{Chupeau:2016jf}
Chupeau M, B{\'e}nichou O, Redner S
\newblock (2016) {Universality classes of foraging with resource renewal}.
\newblock \emph{Phys. Rev. E} 93:032403.

\bibitem{Persson:1998hz}
Persson L, Leonardsson K, De~Roos AM, Gyllenberg M, Christensen B
\newblock (1998) {Ontogenetic scaling of foraging rates and the dynamics of a
  size-structured consumer-resource model}.
\newblock \emph{Theor Popul Biol} 54:270--293.

\bibitem{murray2011mathematical}
Murray JD
\newblock (2011) \emph{{Mathematical Biology: I. An Introduction}},
  Interdisciplinary Applied Mathematics
\newblock (Springer New York) Vol.{} 110.

\bibitem{Strogatz:2008wo}
Strogatz SH
\newblock (2008) \emph{{Nonlinear Dynamics and Chaos: With Applications to
  Physics, Biology, Chemistry, and Engineering}}, Studies in nonlinearity
\newblock (Westview Press, Boulder).

\bibitem{GuckHolmes}
Guckenheimer J, Holmes P
\newblock (1983) \emph{{Nonlinear Oscillations, Dynamical Systems, and
  Bifurcations of Vector Fields}}
\newblock (Springer, New York).

\bibitem{Gross:2004p2428}
Gross T, Feudel U
\newblock (2004) {Analytical search for bifurcation surfaces in parameter
  space}.
\newblock \emph{Physica D} 195:292--302.

\bibitem{Hastings:2001jh}
Hastings A
\newblock (2001) {Transient dynamics and persistence of ecological systems}.
\newblock \emph{Ecol. Lett.} 4:215--220.

\bibitem{Neubert:1997wk}
Neubert M, Caswell H
\newblock (1997) {Alternatives to resilience for measuring the responses of
  ecological systems to perturbations}.
\newblock \emph{Ecology} 78:653--665.

\bibitem{Caswell:2005eo}
Caswell H, Neubert MG
\newblock (2005) {Reactivity and transient dynamics of discrete-time ecological
  systems}.
\newblock \emph{J Differ Equ Appl} 11:295--310.

\bibitem{Neubert:2009td}
Neubert M, Caswell H
\newblock (2009) {Detecting reactivity}.
\newblock \emph{Ecology} 90:2683--2688.

\bibitem{Yodzis:1992hg}
Yodzis P, Innes S
\newblock (1992) {Body size and consumer-resource dynamics}.
\newblock \emph{Am. Nat.} 139:1151--1175.

\bibitem{Brown:2004wq}
Brown J, Gillooly J, Allen A, Savage V, West G
\newblock (2004) {Toward a metabolic theory of ecology}.
\newblock \emph{Ecology} 85:1771--1789.

\bibitem{West:2002it}
West GB, Woodruff WH, Brown JH
\newblock (2002) {Allometric scaling of metabolic rate from molecules and
  mitochondria to cells and mammals.}
\newblock \emph{Proc. Natl. Acad. Sci. USA} 99 Suppl 1:2473--2478.

\bibitem{DeLong:2010dy}
DeLong JP, Okie JG, Moses ME, Sibly RM, Brown JH
\newblock (2010) {Shifts in metabolic scaling, production, and efficiency
  across major evolutionary transitions of life}.
\newblock \emph{Proc. Natl. Acad. Sci. USA} 107:12941--12945.

\bibitem{West:2001bv}
West GB, Brown JH, Enquist BJ
\newblock (2001) {A general model for ontogenetic growth}.
\newblock \emph{Nature} 413:628--631.

\bibitem{moses2008rmo}
Moses ME, {et~al.}
\newblock (2008) {Revisiting a Model of Ontogenetic Growth: Estimating Model
  Parameters from Theory and Data.}
\newblock \emph{Am. Nat.} 171:632--645.

\bibitem{gillooly2002esa}
Gillooly JF, Charnov EL, West GB, Savage VM, Brown JH
\newblock (2002) {Effects of size and temperature on developmental time}.
\newblock \emph{Nature} 417:70--73.

\bibitem{hou}
Hou C, {et~al.}
\newblock (2008) {Energy uptake and allocation during ontogeny}.
\newblock \emph{Science} 322:736--739.

\bibitem{bettencourt}
Bettencourt LMA, Lobo J, Helbing D, Kuhnert C, West GB
\newblock (2007) {Growth, innovation, scaling, and the pace of life in cities}.
\newblock \emph{Proc. Natl. Acad. Sci. USA} 104:7301--7306.

\bibitem{Folland:2008ij}
Folland JP, Mc~Cauley TM, Williams AG
\newblock (2008) {Allometric scaling of strength measurements to body size}.
\newblock \emph{Eur J Appl Physiol} 102:739--745.

\bibitem{Brown:1993p708}
Brown J, Marquet P, Taper M
\newblock (1993) {Evolution of body size: consequences of an energetic
  definition of fitness}.
\newblock \emph{Am. Nat.} 142:573--584.

\bibitem{West:1997cg}
West GB, Brown JH, Enquist BJ
\newblock (1997) {A general model for the origin of allometric scaling laws in
  biology}.
\newblock \emph{Science} 276:122--126.

\bibitem{West:2002ud}
West GB, Woodruff WH, Brown JH
\newblock (2002) {Allometric scaling of metabolic rate from molecules and
  mitochondria to cells and mammals.}
\newblock \emph{Proc. Natl. Acad. Sci. USA} 99 Suppl 1:2473--2478.

\bibitem{Millar:1990p923}
Millar J, Hickling G
\newblock (1990) {Fasting endurance and the evolution of mammalian body size}.
\newblock \emph{Funct. Ecol.} 4:5--12.

\bibitem{Alroy:1998p1594}
Alroy J
\newblock (1998) {Cope's rule and the dynamics of body mass evolution in North
  American fossil mammals}.
\newblock \emph{Science} 280:731--734.

\bibitem{Clauset:2009fh}
Clauset A, Redner S
\newblock (2009) {Evolutionary model of species body mass diversification}.
\newblock \emph{Phys. Rev. Lett.} 102:038103.

\bibitem{Smith:2010p3442}
Smith FA, {et~al.}
\newblock (2010) {The evolution of maximum body size of terrestrial mammals}.
\newblock \emph{Science} 330:1216--1219.

\bibitem{Saarinen:2014br}
Saarinen JJ, {et~al.}
\newblock (2014) {Patterns of maximum body size evolution in Cenozoic land
  mammals: Eco-evolutionary processes and abiotic forcing}.
\newblock \emph{Proc Biol Sci} 281:20132049.

\bibitem{Damuth:1987kr}
Damuth J
\newblock (1987) {Interspecific allometry of population density in mammals and
  other animals: the independence of body mass and population energy-use}.
\newblock \emph{Biol. J. Linn. Soc.} 31:193--246.

\bibitem{DeLong:2012kw}
DeLong JP, Vasseur DA
\newblock (2012) {A dynamic explanation of size{\textendash}density scaling in
  carnivores}.
\newblock \emph{Ecology} 93:470--476.

\bibitem{allen2002}
Allen AP, Brown JH, Gillooly JF
\newblock (2002) {Global biodiversity, biochemical kinetics, and the
  energetic-equivalence rule}.
\newblock \emph{Science} 297:1545--1548.

\bibitem{enquist1998}
Enquist BJ, Brown JH, West GB
\newblock (1998) {Allometric scaling of plant energetics and population
  density}.
\newblock \emph{Nature} 395:163--165.

\bibitem{Kempes:2016}
Kempes CP, Wang L, Amend JP, Doyle J, Hoehler T
\newblock (2016) {Evolutionary tradeoffs in cellular composition across diverse
  bacteria}.
\newblock \emph{ISME J} 10:2145--2157.

\bibitem{tilman1981}
Tilman D
\newblock (1981) {Tests of resource competition theory using four species of
  lake michigan algae}.
\newblock \emph{Ecology} 62:802--815.

\bibitem{dutkiewicz2009}
Dutkiewicz S, Follows MJ, Bragg JG
\newblock (2009) {Modeling the coupling of ocean ecology and biogeochemistry}.
\newblock \emph{Global Biogeochem. Cycles} 23:1--15.

\bibitem{barton2010}
Barton AD, Dutkiewicz S, Flierl G, Bragg J, Follows MJ
\newblock (2010) {Patterns of diversity in marine phytoplankton}.
\newblock \emph{Science} 327:1509--1511.

\end{thebibliography}

\begin{acknowledgments}
  We thank Luis Bettencourt, Jean Philippe Gibert, Eric Libby, and Seth Newsome for helpful
  discussions and comments on the manuscript.  J.D.Y. was supported by
  startup funds at the University of California, Merced, and an Omidyar
  Postdoctoral Fellowship at the Santa Fe Institute.  C.P.K. was supported by
  an Omidyar Postdoctoral Fellowship at the Santa Fe Institute.  S.R. was
  supported by grants DMR-1608211 and 1623243 from the National Science
  Foundation, and by the John Templeton Foundation, all at the Santa Fe
  Institute.
\end{acknowledgments}

%\onecolumn
%\let\cleardoublepage\clearpage

@@

\end{article}

% 
% \begin{figure*}
% \centering
% \includegraphics[width=0.5\textwidth]{fig_Rates.eps}
% \caption{
% Allometrically constrained starvation rate $\sigma=1/t_{\sigma}$ (red), mortality $\mu=1/t_{\mu}$ (blue) and recovery rate $\rho=1/t_{\rho}$ (green) relative to the reproductive rate $\lambda=1/t_{\lambda}$ (black) as a function of body mass (see Equations \ref{t1}, \ref{rhotimescale}, \ref{eq:sigma}, and \ref{mutimescale}).
% The rate of starvation is greater than the rate of reproduction for all realized terrestrial endotherm body sizes.
% Mean values $\pm 25\%$ variation are shown by the shaded region for each rate.
% }
% \label{fig:gvs}
% \end{figure*}

% 
% \begin{figure*}
% \centering
% \includegraphics[width=0.5\textwidth]{fig_DataHopf.eps}
% \caption{ Transcritical (vertical lines) and Hopf bifurcations (curves) for
%   allometrically determined starvation $\sigma$ and recovery $\rho$ rates as
%   a function of different mammalian body sizes: $M=P\times10^2$g (blue) and
%   $M=P\times10^6$g (green), where $P$ is a random uniform variable in $[1,9]$.
%   Points denote realized values of $\sigma$ and $\rho$ given the drawn values for $M$.
%   }
% \label{fig:hopf}
% \end{figure*}

%
%
%	 \begin{figure}[h][h]
%		 \centering
%		 \includegraphics[width=0.5\textwidth]{Growth-trajectory-diagram.pdf}
%		 \caption{
%		 }
%		 \label{growth-diagram}
%	 \end{figure}
%
%	 \begin{figure}[h][h]
%		 \centering
%		 \includegraphics[width=0.25\textwidth]{fig_GvS.pdf}
%		 \caption{
%		 }
%		 \label{GvS}
%	 \end{figure}
%
%
%	\begin{figure}[h][h]
%	 	\centering
%	 	\includegraphics[width=0.25\textwidth]{fig_FixedPoint.pdf}
%	 	\caption{
%	 	}
%		\label{Hopfb}
%	\end{figure}
%
%
%	\begin{figure}[h][h]
%		\centering
%		\includegraphics[width=0.8\textwidth]{fig_DataHopf.pdf}
%		\caption{
%		}
%		\label{DataHopf}
%	\end{figure}
%
%
%	\begin{figure}[h][h]
%		\centering
%	 	\includegraphics[width=0.8\textwidth]{fig_ExtinctionAllometric.pdf}
%	 	\caption{
%	 	}
%	 	\label{Ext}
%	\end{figure}
%
%	\begin{figure}[h][h]
%	 	\centering
%	 	\includegraphics[width=0.5\textwidth]{fig_FPAllometric.pdf}
%	 	\caption{
%	 	}
%	 	\label{Asymp}
%	\end{figure}
%
%\clearpage

>>>>

%% Template for a preprint Letter or Article for submission
%% to the journal Nature.
%% Written by Peter Czoschke, 26 February 2004
%%

\documentclass{nature-2}

\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{subfigure}
\usepackage{tikz}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{array}
\usepackage{tabularx}
\usepackage{multirow}
\usepackage{multicol}
% \usepackage{multibox}
\usepackage{rotating}
\usepackage{color}
\usepackage{layout}
\usepackage{epstopdf}
\usepackage{afterpage}

%% make sure you have the nature.cls and naturemag.bst files where
%% LaTeX can find them

\bibliographystyle{naturemag}

\title{The dynamics of starvation and recovery}

%% Notice placement of commas and superscripts and use of &
%% in the author list

\author{Justin D. Yeakel$^{1,2,3,4}$, Christopher P. Kempes$^{2,3}$ \& Sidney Redner$^{2,3}$}

@@

@@

\begin{affiliations}
 \item School of Natural Science, University of California Merced, Merced, CA, USA
 \item The Santa Fe Institute, Santa Fe, NM, USA
 \item Contributed equally
 \item Corresponding author jdyeakel@gmail.com
\end{affiliations}

\begin{abstract}
The eco-evolutionary dynamics of species are fundamentally linked to the energetic constraints of its constituent individuals. Of particular importance are the tradeoffs between reproduction and the dynamics of starvation and recovery in resource-limited environments. To elucidate the consequences of this tradeoff, we introduce a minimal nutritional state-structured model that incorporates two classes of consumer: nutritionally replete consumers that reproduce, and undernourished, non-reproducing consumers that are susceptible to mortality. As a function of the transition rates between these two states that are determined by the abundance of resources, the consumer populations can either undergo cyclic dynamics or reach a steady state. We obtain strong constraints on starvation and recovery rates by deriving allometric scaling relationships between body size and a variety of traits and find that population dynamics subject to these constraints are typically driven to a steady state. Moreover, we find that these rates fall within a `refuge' in parameter space, where the probability of extinction of the consumer population is minimized. Thus we identify a potential mechanism that may both drive and constrain the dynamics of animal populations. Our model provides a natural framework that predicts maximum body size for mammals by determining the relative stability of an otherwise homogeneous population to a mutant population with altered percentage of body fat. For body masses $< 1.748\times10^7$g, individuals with increased energetic reserves can invade resident populations, and vice versa for body mass $> 1.748\times10^7$g, thus providing a principled mechanism for a within-lineage driver of Cope's rule.
\end{abstract}

%\noindent {\small {\bf Significance Statement} %105/120 words
%Energetic investment in somatic maintenance and growth vs. reproduction directly impacts the dynamics of populations among species. Here, we construct a Nutritional State-structured Model (NSM) to assess the population-level effects of starvation and recovery of a consumer population in a resource-limited environment, and use allometric scaling relationships for mammals to establish all timescales and rates. Our model: i. reveals that mammalian energetic rates minimize the probability of stochastic extinction, ii. establishes dynamic bounds on mammalian body size while providing independent theoretical support for the energy equivalence hypothesis, and iii. provides a mechanistic driver for the evolutionary trend towards larger body size known as Cope's rule.}

@@

@@

@@

@@

@@

@@

\bibliography{aa_starving3}

% \bibliography{ADD}

\begin{addendum}
 \item We thank Luis Bettencourt, Jean Philippe Gibert, Eric Libby, and Seth Newsome for helpful
  discussions and comments on the manuscript.  J.D.Y. was supported by
  startup funds at the University of California, Merced, and an Omidyar
  Postdoctoral Fellowship at the Santa Fe Institute.  C.P.K. was supported by
  an Omidyar Postdoctoral Fellowship at the Santa Fe Institute.  S.R. was
  supported by grants DMR-1608211 and 1623243 from the National Science
  Foundation, and by the John Templeton Foundation, all at the Santa Fe
  Institute.
 \item[Competing Interests] The authors declare that they have no
competing financial interests.
 \item[Correspondence] Correspondence and requests for materials
should be addressed to J.D.Y.~(email:  jdyeakel@gmail.com).
\end{addendum}

\clearpage
\newpage

%FIGURES

\renewcommand{\figurename}{{\bf Figure}}
\renewcommand{\thefigure}{{\bf \arabic{figure}}}

@@

\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{fig_FixedPoint2.eps}
\caption{\small{ The transcritical (dashed) and Hopf bifurcation (solid) as a
  function of the starvation rate $\sigma$ and recovery rate $\rho$ for a 100g consumer.  These
  bifurcation conditions separate parameter space into unphysical, cyclic,
  and steady state dynamic regimes.  The colors show the steady state densities for the energetically replete consumers $F^*$, (warmer colors denote higher densities).  
  % Steady state densities for the energetically deficient consumers $H^*$ (not shown)
  % scale with those for $F^*$.  
}\label{fig:fp}}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{Growth-trajectory-diagram.eps}
\caption{\small{ The growth trajectory over absolute time of an individual organism as a function of body mass.  
Initial growth follows the black trajectory to an energetically replete reproductive adult mass $m=\epsilon_\lambda M$. %which we assume is 95\% asymptotic mass $M$.  
Starvation follows the red trajectory to $m = \epsilon_\sigma \epsilon_\lambda  M$. 
Recovery follows the green curve to the replete adult mass, where this trajectory differs from the original growth because only fat is being regrown which requires a longer time to reach $\epsilon_\lambda M$. %different energetics and 
Alternatively, death from starvation follows the blue trajectory to $m=\epsilon_\mu \epsilon_\lambda  M$.}\label{fig:growth}}
\end{figure}

\begin{figure*}
\centering
\includegraphics[width=0.9\textwidth]{fig_ExtinctionAllometricComb3.eps}
\caption{\small{ Probability of extinction for a consumer with (\emph{A}) $M=10^2$g, (\emph{B}) $M=10^4$g, and (\emph{C}) $M=10^6$g as a function of the starvation rate $\sigma$ and recovery rate $\rho$, where the initial density is given as $(XF^*,XH^*,R^*)$, where $X$ is a random uniform variable in $[0,2]$. Note the change in scale for $M=10^4$ and $M=10^6$g.  Extinction is defined as the population trajectory falling below $0.2\times$ the allometrically constrained steady state. The white points denote the allometrically constrained starvation and recovery rate.}\label{fig:ext}}
\end{figure*}

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{fig_FPAllometric2.eps}
\caption{\small{ (\emph{A}) Consumer steady states $F^*$ (green) and $H^*$ (orange) as a function of
  body mass. Inset: Resource steady state $R^*$ as a function of consumer body mass.
  (\emph{B}) Total energetic use $B_{\rm tot}$ of consumer populations at the steady state as a function of body mass.
  The data are from Damuth \cite{Damuth:1987kr} and have been converted to total population metabolism using the allometric relationships for metabolic rate (please see SI and Refs. \cite{West:2001bv,hou,moses2008rmo}).}\label{fig:mass}}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=0.6\textwidth]{fig_Invasion.eps}
\caption{\small{ Invasion feasibility for organisms with a proportional change in
  mass $\chi$ against a population with a resident body mass $M$.  The blue
  region denotes proportions of modified mass $\chi$ resulting in successful invasion.  The
  red region denotes values of $\chi$ that result in a mass that is below the
  starvation threshold and are thus infeasible.
  Arrows point to the predicted optimal mass from our model $M_{\rm opt}=1.748\times 10^7$, which may serve as an evolutionary attractor for body mass.
  The black wedge points to the largest body mass known for terrestrial mammals (\emph{Deinotherium} spp.) at $1.74\times10^7$g~\cite{Smith:2010p3442}.}\label{fig:invasion}}
\end{figure}

%\pagestyle{empty}

% EXTENDED FIGURES

%\setcounter{figure}{0}
%\renewcommand{\figurename}{{\bf Extended Figure}}
%\renewcommand{\thefigure}{{\bf \arabic{figure}}}
